Left-linear rules in adhesive categories

### 2024-02-13 icalp/fusione.tex
minima conclusioni

As an initial step, % in performing this exploration, 
we observe that for
left-linear rules a number of fundamental properties that play a role
in the concurrency theory of linear rewriting need to be deeply revised, or
plainly do not hold.
%
Firstly, the notion of sequential independence and Church Rosser
theorem, ensuring that sequential independent steps can be performed
in reverse order, does not hold out of the box. Even worse, given two
independent steps it might be possible to switch them in several
different ways. As we will observe, this might prevent the development
of a sensible theory of concurrency as switches are used as a basis
for equating sequences of rewriting steps that differs only for the order of independent
steps.

We single out a class of left-linear rewriting
systems, referred to as \emph{well-switching rewriting systems}, where sequential independence ensures existence and uniqueness of the
switch, and argue that this is the right setting for dealing with
left-linear rewriting systems.

@@

We perform an in-depth investigation of notion of independence between
computational steps in the setting of left-linear rewriting systems
over adhesive categories, which encompasses rewriting over most of the
structures used in semantic and modelling applications.

We showed that the canonical notion of independence adopted for linear
systems does not enjoy some properties which are essential for
developing a sensible theory of concurrency, notably Church Rosser
theorem, and we identified the subclass of well switchable rewriting
systems as an appropriate setting where instead a number of vital
results can be re-established. Specifically, Church Rosser theorem,
i.e., the possibility of performing independent steps in any order is
fully recovered. Moreover, the switch construction and correspondingly
switch equivalence which is at the core of a theory of concurrency for
adhesive recwritng suystems enjoy a number of fundamental properties:
confluence of switching, a weak form of globality of independence and
the existence of a canonical form for switch equivalence proofs.

The class of well-switchabke adhesive rewting systems is vast. Generalising a result in~\cite{baldan2011adhesivity} we can show that one of its defining properties (the fact that sequential independence implies switchability) holds for a class of adhesive categories which include all presheaves over set. Moreover, the second property, uniqueness of switching, is ensured when rewriting possibly hierarchical graphical structures, as long as elements of the top sorts are not merged (e.g., one can consider rewriting systems over directed graphs where only nodes are merged, as it happens in most approaches to modelling calculi, biological systems and also in string diagrams).\todo{Giusto, referneces?}

>>>>

As an initial step, % in performing this exploration, 
we observe that for
left-linear rules a number of fundamental properties that play a role
in the concurrency theory of linear rewriting need to be deeply revised, or
plainly do not hold.
%
Firstly, the notion of sequential independence and Church-Rosser
theorem, ensuring that sequential independent steps can be performed
in reverse order, does not hold out of the box. Even worse, given two
independent steps it might be possible to switch them in several
different ways. As we will observe, this might prevent the development
of a sensible theory of concurrency as switches are used as a basis
for equating sequences of rewriting steps that differs only for the order of independent
steps.

We single out a class of left-linear rewriting
systems, referred to as \emph{well-switchable rewriting systems}, 
where sequential independence ensures existence and uniqueness of the
switch, and argue that this is the right setting for dealing with
left-linear rewriting systems.

@@

We performed an in-depth investigation of notion of independence between
rewriting steps in the setting of left-linear rewriting systems
over adhesive categories, which encompasses most of the
structures commonly used in analysis and modelling applications.

We showed that the canonical notion of independence adopted for linear
systems does not enjoy some properties that are essential for
developing a sensible theory of concurrency, notably a Church-Rosser
theorem, and we identified a subclass of left-linear rewriting systems,
which we call well-switchable, as an appropriate setting where 
many vital results can be re-established. Specifically, a Church-Rosser 
theorem, i.e. the possibility of performing independent steps in any order, is
fully recovered. Moreover, the switch construction and correspondingly
the switch equivalence, at the core of a theory of concurrency for
rewriting systems, enjoy a number of fundamental properties:
confluence of switching, a weak form of globality of independence and
the existence of a canonical form for switch equivalence proofs.

The class of well-switchable adhesive rewriting systems is vast. Generalising a result in~\cite{baldan2011adhesivity} we can show that one of its defining properties (the fact that sequential independence implies switchability) holds for a class of adhesive categories which include all presheaves over set. Moreover, the second property, uniqueness of switching, is ensured when rewriting possibly hierarchical graphical structures, as long as elements of the top sorts are not merged (e.g., one can consider rewriting systems over directed graphs where only nodes are merged, as it happens in most approaches to modelling calculi, biological systems and also in string diagrams).\todo{Giusto, referneces?}
